Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2014-12-21 papers/wdmerger_I/paper.tex
Move all wdmerger_I paper files into the papers/wdmerger_I subdirectory.
Addition of a numerical test section on Galilean invariance, as well as a proper introduction to the numerical test section itself.

A related issue that required a code improvement was that in cases of large density gradients such as the edge of a star, it is possible to generate negative densities in zones even with the more strongly limited PPM. This can occur if a region of large density is moving away from an ambient zone at relatively large speeds; then the net density flux in the ambient zones can be large enough to unphysically drag the density below zero. In practice, this will occur at the trailing edge of a star that is moving across a grid. In such a situation, there are two main approaches one could take: either explicitly introduce a positivity-guaranteeing diffusive flux, or reset the characteristics of the affected zone. We choose the latter approach. Even though it is non-conservative, it preserves a characteristic we value, which is to keep the edge of the stars relatively sharp, as they physically should be. Since the mass of the affected zones is typically already fairly low, this should not seriously affect the energy conservation properties of our simulation. Similarly, it is also possible in this physical situation for very large velocities to be generated in the material behind the star. The momentum transferred to the ambient zone is close to the momentum of the stellar material, but this is of much higher density and so the velocity of the ambient material must become very large to compensate. While this may be physically viable in certain circumstances, it poses a problem for our code, which uses subcycling in time (see below). The timestep is fixed at the beginning of the set of subcycled timesteps, and if in the middle of the cycle a velocity is generated that is so large that it violates the CFL criterion for this timestep, we cannot guarantee the stability of the resulting solution since we cannot alter the timestep. (Nor would we want to, because in some cases because the velocity becomes spuriously large and would drag down the global timestep.) Therefore we also insert a check at the end of every hydrodynamics timestep that resets the velocity and/or sound-speed (indirectly, using the temperature) of any zone that violates the CFL criterion given the timestep we chose at the beginning of the cycle. In practice, we find that this is mainly necessary only for simulations in the inertial reference frame of the white dwarf merger, which we do not normally do for the reasons stated in Section \ref{Sec:Kepler}.

@@

For the evolution of binary systems, it is most natural to evolve the two stars in a frame that is co-rotating at the same period as the orbital period. CASTRO has the ability to evolve systems in a rotating reference frame, as discussed in the original code paper. Source terms corresponding to the Coriolis and centrifugal force terms are added to the momentum and energy equations. In this frame, the stars essentially remain stationary in their original positions due to the centrifugal force supporting against the gravitational attraction, and will remain this way as long as significant mass transfer does not occur. \cite{swesty00} demonstrated (in the context of neutron star mergers) that conservation of angular momentum is much easier to obtain in the rotating reference frame than in an inertial frame in which stars advect large amounts of material around the domain. We wish to emphasize that although it is commonly stated in the literature that grid-based codes poorly conserve angular momentum, this is not generally true. When the resolution is sufficiently high and a rotating reference frame is employed, excellent conservation properties can result, as demonstrated in Section \ref{Sec:Kepler}. We note that as the stars begin to coalesce, the rotating reference frame will no longer provide a good approximation to the spatial motion of the stars and then they will begin to significantly move around the domain. This is not necessarily problematic because the most important feature of the rotating frame is that it helps ensure that the initial coalescence is not the result of spurious numerical loss of angular momentum. When significant mass transfer sets in and evolution proceeds on a dynamical timescale, the conservation properties may be slightly worse but angular momentum conservation is also less important.

@@

\subsection{Gravity}

@@

%==========================================================================
% Test Problems
%==========================================================================
\section{Test Problems}\label{sec:Tests}

In this section we describe a series of test problems that couple the hydrodynamics, gravity, and EOS modules.

\subsection{Maintaining Hydrostatic Equilibrium}\label{Sec:HSE}

@@

\subsection{Gravitational Free Fall}\label{Sec:Gravitational Free Fall}

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}\MarginPar{This result is derived in the freefall directory.}
This result can be derived by recognizing that
\[
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\]
and inserting the velocity as a function of radial separation,
\[
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\]
We determine the initial separation to be consistent with the
simulation performed in Section \ref{Sec:Kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of 100 s for this simulation, so that the separation is
\[
  r_0 = 3.61 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

The analytical result in Equation \ref{analyticalFreeFall} determines
the total elapsed free-fall time,
\[
  t_{\text{ff}} = \frac{\pi}{2} \sqrt{\frac{r_0^3}{2GM}}.
\]
The physical radius of each white dwarf is roughly $10\%$ of the
initial separation, so we consider the evolution terminated when the
radial separation reaches that value (at that point, the stars will no
longer be in free-fall due to physical contact). Since the elapsed
time goes roughly as the square root of the distance for small $r$, we
consider the evolution terminated when $t = 0.99\, t_{\text{ff}}$. The
results of our simulation are shown in Figure \ref{Fig:Free Fall}. The
positions of the two stars are determined by calculating the center of
mass of the right $(x > 0)$ and left $(x < 0)$ sides of the domain at 
the end of each time step, and treating the centers of mass as the 
respective location of the two stars.

\subsection{Keplerian Orbit}\label{Sec:Kepler}

@@

effect of boundary conditions (size of domain and James vs.\ double monopole)

>>>>

A related issue that required a code improvement was that in cases of large density gradients such as the edge of a star, it is possible to generate negative densities in zones even with the more strongly limited PPM. This can occur if a region of large density is moving away from an ambient zone at relatively large speeds; then the net density flux in the ambient zones can be large enough to unphysically drag the density below zero. In practice, this will occur at the trailing edge of a star that is moving across a grid. In such a situation, there are two main approaches one could take: either explicitly introduce a positivity-guaranteeing diffusive flux, or reset the characteristics of the affected zone. We choose the latter approach. Even though it is non-conservative, it preserves a characteristic we value, which is to keep the edge of the stars relatively sharp, as they physically should be. Since the mass of the affected zones is typically already fairly low, this should not seriously affect the energy conservation properties of our simulation. Similarly, it is also possible in this physical situation for very large velocities to be generated in the material behind the star. The momentum transferred to the ambient zone is close to the momentum of the stellar material, but this is of much higher density and so the velocity of the ambient material must become very large to compensate. While this may be physically viable in certain circumstances, it poses a problem for our code, which uses subcycling in time (see below). The timestep is fixed at the beginning of the set of subcycled timesteps, and if in the middle of the cycle a velocity is generated that is so large that it violates the CFL criterion for this timestep, we cannot guarantee the stability of the resulting solution since we cannot alter the timestep. (Nor would we want to, because in some cases because the velocity becomes spuriously large and would drag down the global timestep.) Therefore we also insert a check at the end of every hydrodynamics timestep that resets the velocity and/or sound-speed (indirectly, using the temperature) of any zone that violates the CFL criterion given the timestep we chose at the beginning of the cycle. In practice, we find that this is mainly necessary only for simulations in the inertial reference frame of the white dwarf merger, which we do not normally do for the reasons stated in Section \ref{sec:Kepler}.

@@

For the evolution of binary systems, it is most natural to evolve the two stars in a frame that is co-rotating at the same period as the orbital period. CASTRO has the ability to evolve systems in a rotating reference frame, as discussed in the original code paper. Source terms corresponding to the Coriolis and centrifugal force terms are added to the momentum and energy equations. In this frame, the stars essentially remain stationary in their original positions due to the centrifugal force supporting against the gravitational attraction, and will remain this way as long as significant mass transfer does not occur. \cite{swesty00} demonstrated (in the context of neutron star mergers) that conservation of angular momentum is much easier to obtain in the rotating reference frame than in an inertial frame in which stars advect large amounts of material around the domain. We wish to emphasize that although it is commonly stated in the literature that grid-based codes poorly conserve angular momentum, this is not generally true. When the resolution is sufficiently high and a rotating reference frame is employed, excellent conservation properties can result, as demonstrated in Section \ref{sec:Kepler}. We note that as the stars begin to coalesce, the rotating reference frame will no longer provide a good approximation to the spatial motion of the stars and then they will begin to significantly move around the domain. This is not necessarily problematic because the most important feature of the rotating frame is that it helps ensure that the initial coalescence is not the result of spurious numerical loss of angular momentum. When significant mass transfer sets in and evolution proceeds on a dynamical timescale, the conservation properties may be slightly worse but angular momentum conservation is also less important.

@@

\subsection{Gravity}
\label{sec:gravity}

@@

%==========================================================================
% Test Problems
%==========================================================================
\section{Numerical Test Problems}\label{sec:Tests}

Merger simulations face a number of numerical difficulties that are not present in single-degenerate Type Ia and core-collapse supernova simulations. In Section \ref{sec:gravity}, we discussed how the lack of spherical symmetry necessitates a careful look at the gravity solver. There are also hydrodynamical issues: the merger process will result in substantial motion of stellar material across the grid. This bulk motion presents an opportunity for advection errors to build up, and is only partially mitigated by evolving the white dwarfs in a co-rotating frame. It is therefore important to be be aware of the behavior of the code in such circumstances. The behavior of CASTRO for many standard hydrodynamics test problems was detailed in the original code paper \citep{castro}, and in the interest of brevity we will not repeat them here. Instead, we focus on a subset of problems that highlight the special difficulties introduced in merger simulations. These problems couple the hydrodynamics, gravity and equation of state modules. We observe that while in most non-trivial three-dimensional problems this creates a complexity that makes it impossible to determine exact analytical solutions, it is straightforward to devise problems for which certain global properties should obey simple, expected behaviors. Where possible, these should be quantified and a convergence study performed. We recommend that all simulation codes should check their performance on such problems to ensure their reliability when coupling the various physics modules necessary to build a complete and realistic simulation.

\subsection{Maintaining Hydrostatic Equilibrium}\label{sec:HSE}

@@

This problem was studied in the first CASTRO paper, but is worth revisiting here. A single star explosion simulation may only last a couple of seconds, and the CASTRO paper studied the behavior of the star after one second of evolution. However, the dynamical timescale of a typical carbon-oxygen white dwarf is on the order of 1--10 seconds. Additionally, a binary orbit is typically on the order of 10--100 seconds when a merger simulation starts, and with equilibrium initial conditions the system may survive for tens of orbits before the secondary is disrupted. When this does happen, we want to be confident that it was because of the dynamics of the merger process and not because of an instability in an individual star. Our goal here is thus to populate a single star onto our three-dimensional coordinate grid and evolve it for a period of time long enough to assess whether the star is truly stable, and to probe how the size of deviation from equilibrium is affected by grid resolution.

\subsection{Gravitational Free Fall}\label{sec:Gravitational Free Fall}

A simple test to verify the Poisson solver implemented by CASTRO is
the case of gravitational free fall. In this setup, two stars, each
individually in an equilibrium state, are placed on the computational
grid, separated by an initial distance $r_0$ along the $x$ axis with
zero initial velocity. We choose stars of masses $0.6$ and $0.8\,
M_\odot$, with the lower mass star on the left side (e.g. $x < 0$) of
the domain and the higher mass star on the right, such that their
center of mass coincides with the center of the
domain. Gravitationally, the stars may be treated as point masses
until the point of contact, so the equation of motion governing the
distance $r$ between their centers of mass is that of simple free
fall:
\begin{equation}
  \ddot{r}(t) = - \frac{GM}{r},
\end{equation}
where $G$ is the gravitational constant and $M$ is the total mass of
the system. This differential equation has a closed-form solution for
the evolution time as a function of separation:
\begin{equation}
  t(r) = \sqrt{\frac{r_0^3}{2GM}} \left[ \text{arccos}\left(\sqrt{\frac{r}{r_0}}\,\right) + \sqrt{\frac{r}{r_0} \left(1 - \frac{r}{r_0}\right)}\ \right]. \label{analyticalFreeFall}
\end{equation}\MarginPar{This result is derived in the freefall directory.}
This result can be derived by recognizing that
\[
  t(r) = \int_{r_0}^r \frac{dr}{v(r)}
\]
and inserting the velocity as a function of radial separation,
\[
  v(r) = \sqrt{\frac{2GM}{r_0}\left(\frac{r_0}{r} - 1\right)}.
\]
We determine the initial separation to be consistent with the
simulation performed in Section \ref{sec:Kepler}; that is, we select
an initial orbital period and use Kepler's third law to calculate the
radius of a circular orbit with that period. We select an initial
orbital period of $T = 100$ s for this simulation, so that the separation is
\[
  r_0 = 3.61 \times 10^{9}\ \text{cm}.
\]
We chose a relatively low resolution simulation to demonstrate the
capabilities of CASTRO even while using modest resources. The
computational grid is covered by a coarse grid of $48^3$ zones, with
two levels of refinement above the coarse grid. Each refined grid
carries an increase in resolution by a factor of 4 relative to the
coarser grid below it. CASTRO initially assigns $93\%$ of the domain
to be covered by the intermediate resolution grids, and $0.05\%$ of
the domain to be covered by the finest resolution grids.

The analytical result in Equation \ref{analyticalFreeFall} determines
the total elapsed free-fall time,
\[
  t_{\text{ff}} = \frac{\pi}{2} \sqrt{\frac{r_0^3}{2GM}} = \frac{T}{4\sqrt{2}}.
\]
The physical radius of each white dwarf is roughly $10\%$ of the
initial separation, so we consider the evolution terminated when the
radial separation reaches that value (at that point, the stars will no
longer be in free-fall due to physical contact). Since the elapsed
time goes roughly as the square root of the distance for small $r$, we
consider the evolution terminated when $t = 0.99\, t_{\text{ff}}$. The
results of our simulation are shown in Figure \ref{Fig:Free Fall}. The
positions of the two stars are determined by calculating the center of
mass of the right $(x > 0)$ and left $(x < 0)$ sides of the domain at 
the end of each time step, and treating the centers of mass as the 
respective location of the two stars.

\subsection{Galilean Invariance}\label{sec:galileo}

It is often stated in the literature that Eulerian methods for hydrodynamics with grids fixed in space do not obey the Galilean invariance of the underlying Euler equations, so that simulations moving at a uniform bulk velocity will appear different than an equivalent stationary simulation. If true, this means that codes such as CASTRO may be fundamentally inappropriate for merger problems (though any errors should be greatly diminished in a reference frame that rotates with the initial binary orbit). Recently, this has come up in two ways which are of note for us in the present study. We will explain these situations and run tests to determine whether this actually is a significant concern.

\citet{arepo} performed a Kelvin-Helmholtz instability test and showed that (at low resolution) a fixed-grid code failed to develop the expected fluid instability when the whole fluid was moving at a strongly supersonic uniform velocity. This contrasted with the results of the moving-mesh code AREPO being presented in that study, which demonstrated Galilean invariance even at large bulk velocities. If verified, this claim would have important consequences on how much we can trust the ability of CASTRO to test the violent merger progenitor model. Shearing between the material flowing out of the secondary and material near the surface of the primary may trigger fluid instabilities that play an important role in the evolution of that gas, which is the site of the initial detonation in the prompt explosion model. \citet{guillochon:2010} showed for their simulation that Kelvin-Helmholtz instabilities produced this way may raise the temperature of the accreting material enough to ignite a detonation. Therefore if we are not correctly reproducing the characteristics of the Kelvin-Helmholtz instability in the case where there is significant mass motion on the grid, we cannot be confident that a detonation (or lack thereof) is not numerically seeded. However, \citet{robertson:2010} observe that Galilean invariance of simulation results for the Euler equations occurs only because of truncation error in the discretization of the fluid equations. This takes the form of a numerical diffusion term which is dependent on velocity (and also resolution). The advantage of a moving-mesh code is that the mesh everywhere moves with the local flow velocity, which substantially reduces the numerical diffusion. \citeauthor{robertson:2010} argue that the differences seen between the moving-mesh and fixed-grid code are caused by the interaction of this numerical diffusion with small-scale instabilities (which may be physical or numerical) which couple with and fundamentally alter the large-scale modes. Small-scale instabilities are seeded in this problem by the choice of a sharp initial discontinuity between the fluids. Crucially though, \citeauthor{robertson:2010} point out that this problem does not converge with resolution and so it is not possible to know the correct behavior of this problem. As such, we do not know whether the small-scale modes found in AREPO are real, so the problem is not useful in formally discriminating between methodologies. They instead propose an alternate test with a smoother initial contact. This converges to the same solution qualitatively in both the stationary and bulk velocity cases, indicating that the code does generally maintain Galilean invariance (to some specified error that depends on resolution and the uniform flow speed).  We will see whether we can reproduce this result, as well as a result by \cite{wadsley:2008}, who used the FLASH code to simulate a hot bubble subject to mixing by the Kelvin-Helmholtz instability, and also found that the mixing was affected by a uniform bulk velocity.

A related question is whether our code reliably simulates the bulk motion of the stars across the grid, and whether such bulk motion affects the stability of the star. This concern is prompted by the study of \cite{tasker:2008}, who studied the effect of uniform translation on the stability of a spherically symmetric model for a galaxy cluster. They compared the radial profile of the cluster at initialization and after a period of time evolution. Using FLASH and ENZO, they found that a static cluster retains its shape at high enough resolution, while uniform translation of the cluster causes mixing of the core material due to numerical diffusion which results in an underestimation of the core's true density. The SPH codes they used did a better job maintaining the core density. We will perform a variant of this test using white dwarf models.

\subsubsection{Kelvin-Helmholtz Instability}\label{sec:khi}

\subsubsection{Hot Bubble}\label{sec:hot_bubble}

\subsubsection{Moving Star}\label{sed:moving_star}

\subsection{Keplerian Orbit}\label{sec:Kepler}

@@

effect of boundary conditions (size of domain and possibly multipole dependence)
